Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-04-28 joint_inf.tex
add appendix; address most issues

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \argmax_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
We can perform this by computing the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t),
\end{equation}
then estimating the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\fullAlignment;\tau, t)
\end{equation}
while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.

@@

L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\

@@

where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.

@@

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments can be described as
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips at which a character is assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
$$
\mathcal{P}(\siteSplitSet) = \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
$$
Had we not identified character complements, we would need to consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
%EM Would it be worth enumerating the site splits ahead of time? It seems a little funky to be calling them j each time.
The probabilities of observing ancestral splits for $\siteSplit_j=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
Thus the set of most likely ancestral states contains a single element in this case, here $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_j\in\ancestralSplitPartition_j(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_j=\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2.
$$
In this case, the set of most likely ancestral states is $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_j\in\ancestralSplitPartition_j(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3.
$$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
%EM As described IRL, a little strange to be redefining (9) with an extra variable. After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}.
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
%EM did you mean to drop the eta from the first term of this equation?
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
%EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
%EM Suggest \hat{\fullAncestralSplitPartitions}_1, etc.
\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau',t_1),\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau^*,t_2),\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}(\tau',t_1)\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}(\tau^*,t_2)\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any potential example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
%EM Seems a little strange to give a new equation number to an equation that already exists with an assigned number.
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t})
$$
using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\hat{\fullAncestralSplitPartitions}(\tau,t)\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t}).
$$

% vvvv Remove these? I think they're obvious now, but maybe they can be made more explicit above?
%EM I agree-- quite clear now!
\textbf{Argument for lower bound}: This makes intuitive sense, though may be difficult to show in general.
Can we not just say $\max_u f(u) + g(u)$ is greater than $f(u') + g(u')$ for any $u'$ that doesn't maximize $f(u) + g(u)$ just by definition?
There could be multiple maxima, but that's why we have greater than or equal to.
I think the only thing we'd need to show is that a maximum exists, but since $\{x,y,w\}$ are supported on a closed set this is certainly true.
I must be missing something.

\textbf{Argument for upper bound}: I think this is true mostly because $\max_u f(u) + g(u)$ should be less than $\max_u f(u) + \max_u g(u)$ by some triangle inequality argument.
If we were to actually maximize the full likelihood, we should get something smaller than maximizing some form of it that we consider by looking at the parts separately.
For a fixed $\{x^*, y^*\}$ we can calculate this upper bound directly---can we obtain it as a useful-looking function of these parameters in the general case?

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
%EM clarify that we aren't assuming that the x and y in t are the same as those appearing in t^*?
Call $t=\{x,y,w\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.

We will parameterize these trees \textbf{not with branch lengths}, but rather such that the probability of a substitution on a branch with parameter $x$ is $p_x=1-2x$.
%EM (Scratching head) isn't p_x=1-2x backwards? Then p_x will take values in the range -1 to 1.
Here, $\{p_x,p_y,p_w\}\in[0,1/2]^3$ and $\{x,y,w\}\in[0,1]^3$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have nice algebraic properties, as can be seen via the Hadamard transform.

\textbf{Justification for estimand:} It might be worth arguing why this form of the estimand $t$ is necessary.
%EM definitely. This is a big TODO.
I think the idea is that for either of these topologies, any maximum obtained will occur where the top two and bottom two branches are equal, though not necessarily the middle branch.
This would be good to show explicitly.

@@

\subsubsection{Upper bound for Farris partial likelihood}

@@

    &\qquad := L^{p,1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).

@@

\paragraph{Examples}

See Table~\ref{tab:examples} for parameters, bounds and true likelihood values obtained via basin hopping.

Set $\{x, y\} = \{0.25, 0.25\}$.
For a Farris zone generating topology, we have
$$
L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157,
$$
$$
L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157
$$
and
$$
L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157.
$$
Computing the upper bound yields
$$
\max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t) \le -4.22274479428471
$$
and it remains to show that
$$
\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \ge -4.22274479428471.
$$

>>>>

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
%das: I forgot that standard likelihoods are written with the parameters on the left.
%Since H is like a hidden variable, I went for an expectation-maximization notation, grouping them with the observations.
%It seems a little wonky, though, when we start having maximum likelihood ancestral states depending on \tau,t.
%Oh well...
L_\nCols(\tau, t; \fullAlignment,\fullAncestralStates) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \argmax_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\tau, t; \fullAlignment).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
We can perform this by computing the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates),
\end{equation}
then estimating the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment)
\end{equation}
while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.

@@

L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\

@@

where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

@@

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments can be described as
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips at which a character is assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Had we not identified character complements, we would need to consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
%EM Would it be worth enumerating the site splits ahead of time? It seems a little funky to be calling them j each time.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
Thus the set of most likely ancestral states contains a single element in this case, here $\ancestralSplitPartition_1(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_6=\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2.
$$
In this case, the set of most likely ancestral states is $\ancestralSplitPartition_6(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_6\in\ancestralSplitPartition_6(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_6 \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3.
$$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
%EM As described IRL, a little strange to be redefining (9) with an extra variable. After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t))
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
%EM did you mean to drop the eta from the first term of this equation?
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
%EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
\tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
%EM Suggest \hat{\fullAncestralSplitPartitions}_1, etc.
\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}_1,\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}_2,\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any potential example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
We obtain an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+ \tilde{\ell}_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions},\tau,\hat{t})
$$
using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\hat{\fullAncestralSplitPartitions}\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
+ \tilde{\ell}_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions},\tau,\hat{t}).
$$

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
%EM clarify that we aren't assuming that the x and y in t are the same as those appearing in t^*?
Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.

We will parameterize these trees \textbf{not with branch lengths}, but rather such that the probability of a substitution on a branch with parameter $x'$ is $p_x'=(1-x')/2$.
%EM (Scratching head) isn't p_x=1-2x backwards? Then p_x will take values in the range -1 to 1.
Here, $\{p_x',p_y',p_w'\}\in[0,1/2]^3$ and $\{x',y',w'\}\in[0,1]^3$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have nice algebraic properties, as can be seen via the Hadamard transform.

%das: did this on paper, but can write it out.
\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\{x_1,y_1,x_2,y_2,w\}$ counting from the top left.
The likelihood $\ell(x_1,y_1,x_2,y_2,w)$ can be written as a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
In this case, $x_1$ and $x_2$ can be exchanged for one another without changing the value of $\ell$, and any maximum will require $x_1=x_2$.
A similar argument can be made for $y_1$ and $y_2$.
For the Felsenstein topolgy this is not the case, but, since we are interested in a lower bound for this topology, we will simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

\subsubsection{Upper bound for Farris likelihood}

@@

    &\qquad := L^{1}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).

@@
